---
title: "Q2 - Deaths rates during waves"
format: html
editor: visual
---

```{r}
library(dplyr)
library(tidyverse)
library(lubridate)
library(ggrepel)
```

```{r}
dat <- readRDS('../data/dat.rds')
```

## Question 2: For each period compute the deaths rates by state. Describe which states did better or worse during the different periods.

This documents makes various plots to see death rates during COVID-19 waves. The following waves are:

-   Wave 1: April 2020 - June 2020

-   Wave 2: October 2020 - March 2021

-   Wave 3: August 2021 - November 2021

-   Wave 4: January 2022 - March 2022

For the final paper, we can pick and choose which states/regions to look at in order to tell a story. The rest of the graphs can go into an appendix.

```{r}
#selcting NY for region 2 from wave 1
unique(dat |> 
  filter(region %in% 2) |> 
  select(state))

unique(dat |> 
  filter(region %in% 10) |> 
  select(state))
```

## Wave 1: April - June 2020 

```{r}
dat |>
  select(date, population, 
         covid_deaths, total_deaths, region,
         region_name, state) |>
  pivot_longer(c(covid_deaths, total_deaths), # create col of type of rate, for facet wrapping
               names_to = 'type', 
               values_to = 'values') |> 
  mutate(rate = (values / population) * 100000,
         region = as.factor(region)) |>
  filter(date >= make_date(2020, 3, 15) & date <= make_date(2020, 6, 1)) |>
  ggplot(aes(x = date, y = rate, color = state)) + 
  geom_line(aes(group = state)) + # creates smooth lines
  facet_grid(type ~ region, scales = 'free_y') + # stack/scale axis
  labs(x = 'Time (Calendar Time)', 
       y = 'Rates (per 100000)',
       title = 'Wave 1: Rates of COVID deaths, & total deaths in U.S.',
       color = 'State') + 
  theme(axis.text.x = element_text(vjust = 0.75, size = 6, angle = 45),
        legend.key.size = unit(0.25, 'cm'))


unique(dat |> 
  filter(region %in% 8) |> 
  select(state))
```

```{r}
for (i in sort(unique(dat$region_name))) {
 plot1 <- dat |>
  select(date, population, 
         covid_deaths, total_deaths, region,
         region_name, state) |>
  pivot_longer(c(covid_deaths, total_deaths), # create col of type of rate, for facet wrapping
               names_to = 'type', 
               values_to = 'values') |> 
  mutate(rate = (values / population) * 100000,
         region = as.factor(region)) |>
   filter(region_name %in% i) |>
   #Wave 1: April-June 2020 (look at one month before april and one month after june?)
  filter(date >= make_date(2020, 3, 15) & date <= make_date(2020, 6, 1)) |>
  ggplot(aes(x = date, y = rate, color = state)) + 
  geom_line(aes(group = state)) + # creates smooth lines
  facet_grid(type ~ region, scales = 'free_y') + # stack/scale axis
  labs(x = 'Time (Calendar Time)', 
       y = 'Rates (per 100000)',
       title = paste('Wave 1: Rates of COVID deaths, & total deaths in U.S.', i),
       color = 'State') + 
  theme(axis.text.x = element_text(vjust = 0.75, size = 6, angle = 45),
        legend.key.size = unit(0.25, 'cm'))
   
  print(plot1)
}

```

Based on Wave 1:

Best State:

Worst State: New Jersey

## Wave 2: October 2020 - March 2021 

```{r}
dat |>
  select(date, population, 
         covid_deaths, total_deaths, region,
         region_name, state) |>
  pivot_longer(c(covid_deaths, total_deaths), # create col of type of rate, for facet wrapping
               names_to = 'type', 
               values_to = 'values') |> 
  mutate(rate = (values / population) * 100000,
         region = as.factor(region)) |>
  filter(date >= make_date(2020, 9, 1) & date <= make_date(2021, 4, 1)) |>
  ggplot(aes(x = date, y = rate, color = state)) + 
  geom_line(aes(group = state)) + # creates smooth lines
  facet_grid(type ~ region, scales = 'free_y') + # stack/scale axis
  labs(x = 'Time (Calendar Time)', 
       y = 'Rates (per 100000)',
       title = 'Wave 2: Rates of COVID deaths, & total deaths in U.S.',
       color = 'State') + 
  theme(axis.text.x = element_text(vjust = 0.75, size = 6, angle = 45),
        legend.key.size = unit(0.25, 'cm'))
```

```{r}
for (i in sort(unique(dat$region_name))) {
 plot2 <- dat |>
  select(date, population, 
         covid_deaths, total_deaths, region,
         region_name, state) |>
  pivot_longer(c(covid_deaths, total_deaths), # create col of type of rate, for facet wrapping
               names_to = 'type', 
               values_to = 'values') |> 
  mutate(rate = (values / population) * 100000,
         region = as.factor(region)) |>
   filter(region_name %in% i) |>
   #Wave 2: October 2020 - March 2021 
  filter(date >= make_date(2020, 9, 1) & date <= make_date(2021, 4, 1)) |>
  ggplot(aes(x = date, y = rate, color = state)) + 
  geom_line(aes(group = state)) + # creates smooth lines
  facet_grid(type ~ region, scales = 'free_y') + # stack/scale axis
  labs(x = 'Time (Calendar Time)', 
       y = 'Rates (per 100000)',
       title = paste('Wave 2: Rates of COVID deaths, & total deaths in U.S.', i),
       color = 'State') + 
  theme(axis.text.x = element_text(vjust = 0.75, size = 6, angle = 45),
        legend.key.size = unit(0.25, 'cm'))
   
  print(plot2)
}
```

Based on Wave 2:

Best State:

Worst State: SD (South Dakota)

## Wave 3: August-November 2021 

```{r}
dat |>
  select(date, population, 
         covid_deaths, total_deaths, region,
         region_name, state) |>
  pivot_longer(c(covid_deaths, total_deaths), # create col of type of rate, for facet wrapping
               names_to = 'type', 
               values_to = 'values') |> 
  mutate(rate = (values / population) * 100000,
         region = as.factor(region)) |>
  filter(date >= make_date(2021, 7, 1) & date <= make_date(2021, 12, 31)) |>
  ggplot(aes(x = date, y = rate, color = state)) + 
  geom_line(aes(group = state)) + # creates smooth lines
  facet_grid(type ~ region, scales = 'free_y') + # stack/scale axis
  labs(x = 'Time (Calendar Time)', 
       y = 'Rates (per 100000)',
       title = 'Wave 3: Rates of COVID deaths, & total deaths in U.S.',
       color = 'State') + 
  theme(axis.text.x = element_text(vjust = 0.75, size = 6, angle = 45),
        legend.key.size = unit(0.25, 'cm'))

unique(dat |> 
  filter(region %in% 4) |> 
  select(state))
```

```{r}
for (i in sort(unique(dat$region_name))) {
 plot3 <- dat |>
  select(date, population, 
         covid_deaths, total_deaths, region,
         region_name, state) |>
  pivot_longer(c(covid_deaths, total_deaths), # create col of type of rate, for facet wrapping
               names_to = 'type', 
               values_to = 'values') |> 
  mutate(rate = (values / population) * 100000,
         region = as.factor(region)) |>
   filter(region_name %in% i) |>
   #Wave 2: October 2020 - March 2021 
  filter(date >= make_date(2021, 7, 1) & date <= make_date(2021, 12, 31))  |>
  ggplot(aes(x = date, y = rate, color = state)) + 
  geom_line(aes(group = state)) + # creates smooth lines
  facet_grid(type ~ region, scales = 'free_y') + # stack/scale axis
  labs(x = 'Time (Calendar Time)', 
       y = 'Rates (per 100000)',
       title = paste('Wave 3: Rates of COVID deaths, & total deaths in U.S.', i),
       color = 'State') + 
  theme(axis.text.x = element_text(vjust = 0.75, size = 6, angle = 45),
        legend.key.size = unit(0.25, 'cm'))
   
  print(plot3)
}


```

Based on Wave 3:

Best State:

Worst State: Florida

## Wave 4: January 2022 - March 2022

```{r}
dat |>
  select(date, population, 
         covid_deaths, total_deaths, region,
         region_name, state) |>
  pivot_longer(c(covid_deaths, total_deaths), # create col of type of rate, for facet wrapping
               names_to = 'type', 
               values_to = 'values') |> 
  mutate(rate = (values / population) * 100000,
         region = as.factor(region)) |>
  filter(date >= make_date(2021, 12, 1) & date <= make_date(2022, 3, 31)) |>
  ggplot(aes(x = date, y = rate, color = state)) + 
  geom_line(aes(group = state)) + # creates smooth lines
  facet_grid(type ~ region, scales = 'free_y') + # stack/scale axis
  labs(x = 'Time (Calendar Time)', 
       y = 'Rates (per 100000)',
       title = 'Wave 4: Rates of COVID deaths, & total deaths in U.S.',
       color = 'State') + 
  theme(axis.text.x = element_text(vjust = 0.75, size = 6, angle = 45),
        legend.key.size = unit(0.25, 'cm'))

unique(dat |> 
  filter(region %in% 6) |> 
  select(state))
```

```{r}
for (i in sort(unique(dat$region_name))) {
 plot4 <- dat |>
  select(date, population, 
         covid_deaths, total_deaths, region,
         region_name, state) |>
  pivot_longer(c(covid_deaths, total_deaths), # create col of type of rate, for facet wrapping
               names_to = 'type', 
               values_to = 'values') |> 
  mutate(rate = (values / population) * 100000,
         region = as.factor(region)) |>
   filter(region_name %in% i) |>
   #Wave 2: October 2020 - March 2021 
  filter(date >= make_date(2021, 12, 1) & date <= make_date(2022, 3, 31)) |>
  ggplot(aes(x = date, y = rate, color = state)) + 
  geom_line(aes(group = state)) + # creates smooth lines
  facet_grid(type ~ region, scales = 'free_y') + # stack/scale axis
  labs(x = 'Time (Calendar Time)', 
       y = 'Rates (per 100000)',
       title = paste('Wave 4: Rates of COVID deaths, & total deaths in U.S.', i),
       color = 'State') + 
  theme(axis.text.x = element_text(vjust = 0.75, size = 6, angle = 45),
        legend.key.size = unit(0.25, 'cm'))
   
  print(plot4)
}
```

Based on Wave 4:

Best State:

Worst State: OK
