---
title: "virulence_Q3"
format: html
editor: visual
---

```{r}
library(dplyr)
library(tidyverse)
library(lubridate)
library(ggrepel)
```

```{r}
dat <- readRDS('../data/dat.rds')
```

This documents makes various plots to see virulence during COVID-19 waves. Virulence was analyzed using case rates for each state due to different state populations (i.e., having 5 cases in a week in a population of 10 people is different than having 5 cases in a week in a population of 100000 people). The following waves are:

-   Wave 1: April 2020 - June 2020

-   Wave 2: October 2020 - March 2021

-   Wave 3: August 2021 - November 2021

-   Wave 4: January 2022 - March 2022

For the final paper, we can pick and choose which states/regions to look at in order to tell a story. The rest of the graphs can go into an appendix.

Wave 1: individual states

```{r}
dat |>
  select(date, population, cases,
         covid_deaths, total_deaths, region,
         region_name, state) |>
  filter(date >= make_date(2020, 3, 15) & date <= make_date(2020, 6, 1)) |>
  mutate(cases_rate = (cases / population) * 100000) |>
  ggplot(aes(x = date, y = cases_rate, color = state)) + 
  geom_line(aes(group = state)) + # creates smooth lines
  facet_wrap( ~ region_name, scales = 'free_y') + # stack/scale axis 
  labs(x = 'Time (Calendar Time)', 
       y = 'Cases Rate (per 100000)',
       title = 'Wave 1: COVID case rates in U.S. per state',
       color = 'State') + 
  theme(axis.text.x = element_text(vjust = 0.75, size = 6, angle = 45),
        legend.key.size = unit(0.25, 'cm')) 
```

Wave 1: By region

```{r}
dat |>
  select(date, population, cases,
         covid_deaths, total_deaths, region,
         region_name, state) |>
  filter(date >= make_date(2020, 3, 15) & date <= make_date(2020, 6, 1)) |>
  group_by(date, region_name) |>
  reframe(date = date, 
          population = sum(population),
          cases = sum(cases)) |>
  distinct(.keep_all = TRUE) |>
  mutate(cases_rate = (cases / population) * 100000) |>
  ggplot(aes(x = date, y = cases_rate, color = region_name, group = region_name)) + 
  geom_line() + # creates smooth lines
  labs(x = 'Time (Calendar Time)', 
       y = 'Rates (per 100000)',
       title = 'Wave 1: Total COVID case rates in U.S. per region',
       color = 'Region') + 
  theme(axis.text.x = element_text(vjust = 0.75, size = 6)) 
```

Wave 2: individual states

```{r}
dat |>
  select(date, population, cases,
         covid_deaths, total_deaths, region,
         region_name, state) |>
  filter(date >= make_date(2020, 9, 1) & date <= make_date(2021, 4, 1)) |>
  mutate(cases_rate = (cases / population) * 100000) |>
  ggplot(aes(x = date, y = cases_rate, color = state)) + 
  geom_line(aes(group = state)) + # creates smooth lines
  facet_wrap( ~ region_name, scales = 'free_y') + # stack/scale axis 
  labs(x = 'Time (Calendar Time)', 
       y = 'Cases Rate (per 100000)',
       title = 'Wave 2: COVID case rates in U.S. per state',
       color = 'State') + 
  theme(axis.text.x = element_text(vjust = 0.75, size = 6),
        legend.key.size = unit(0.25, 'cm')) 
```

Wave 2: By region

```{r}
dat |>
  select(date, population, cases,
         covid_deaths, total_deaths, region,
         region_name, state) |>
  filter(date >= make_date(2020, 9, 1) & date <= make_date(2021, 4, 1)) |>
  group_by(date, region_name) |>
  reframe(date = date, 
          population = sum(population),
          cases = sum(cases)) |>
  distinct(.keep_all = TRUE) |>
  mutate(cases_rate = (cases / population) * 100000) |>
  ggplot(aes(x = date, y = cases_rate, color = region_name, group = region_name)) + 
  geom_line() + # creates smooth lines
  labs(x = 'Time (Calendar Time)', 
       y = 'Rates (per 100000)',
       title = 'Wave 2: Total COVID case rates in U.S. per region',
       color = 'Region') + 
  theme(axis.text.x = element_text(vjust = 0.75, size = 6)) 
```

Wave 3: individual states

```{r}
dat |>
  select(date, population, cases,
         covid_deaths, total_deaths, region,
         region_name, state) |>
  filter(date >= make_date(2021, 7, 1) & date <= make_date(2021, 12, 31)) |>
  mutate(cases_rate = (cases / population) * 100000) |>
  ggplot(aes(x = date, y = cases_rate, color = state)) + 
  geom_line(aes(group = state)) + # creates smooth lines
  facet_wrap( ~ region_name, scales = 'free_y') + # stack/scale axis 
  labs(x = 'Time (Calendar Time)', 
       y = 'Cases Rate (per 100000)',
       title = 'Wave 3: COVID case rates in U.S. per state',
       color = 'State') + 
  theme(axis.text.x = element_text(vjust = 0.75, size = 6, angle = 45),
        legend.key.size = unit(0.25, 'cm')) 
```

```{r}
dat |>
  select(date, population, cases,
         covid_deaths, total_deaths, region,
         region_name, state) |>
  filter(date >= make_date(2021, 7, 1) & date <= make_date(2021, 12, 31)) |>
  group_by(date, region_name) |>
  reframe(date = date, 
          population = sum(population),
          cases = sum(cases)) |>
  distinct(.keep_all = TRUE) |>
  mutate(cases_rate = (cases / population) * 100000) |>
  ggplot(aes(x = date, y = cases_rate, color = region_name, group = region_name)) + 
  geom_line() + # creates smooth lines
  labs(x = 'Time (Calendar Time)', 
       y = 'Rates (per 100000)',
       title = 'Wave 3: Total COVID case rates in U.S. per region',
       color = 'Region') + 
  theme(axis.text.x = element_text(vjust = 0.75, size = 6, angle = 45))
```

Wave 4: individual states

```{r}
dat |>
  select(date, population, cases,
         covid_deaths, total_deaths, region,
         region_name, state) |>
  filter(date >= make_date(2021, 12, 1) & date <= make_date(2022, 3, 31)) |>
  mutate(cases_rate = (cases / population) * 100000) |>
  ggplot(aes(x = date, y = cases_rate, color = state)) + 
  geom_line(aes(group = state)) + # creates smooth lines
  facet_wrap( ~ region_name, scales = 'free_y') + # stack/scale axis 
  labs(x = 'Time (Calendar Time)', 
       y = 'Cases Rate (per 100000)',
       title = 'Wave 4: COVID case rates in U.S. per state',
       color = 'State') + 
  theme(axis.text.x = element_text(vjust = 0.75, size = 6, angle = 45),
        legend.key.size = unit(0.25, 'cm')) 
```

Wave 4: by region

```{r}
dat |>
  select(date, population, cases,
         covid_deaths, total_deaths, region,
         region_name, state) |>
  filter(date >= make_date(2021, 12, 1) & date <= make_date(2022, 3, 31)) |>
  group_by(date, region_name) |>
  reframe(date = date, 
          population = sum(population),
          cases = sum(cases)) |>
  distinct(.keep_all = TRUE) |>
  mutate(cases_rate = (cases / population) * 100000) |>
  ggplot(aes(x = date, y = cases_rate, color = region_name, group = region_name)) + 
  geom_line() + # creates smooth lines
  labs(x = 'Time (Calendar Time)', 
       y = 'Rates (per 100000)',
       title = 'Wave 4: Total COVID case rates in U.S. per region',
       color = 'Region') + 
  theme(axis.text.x = element_text(vjust = 0.75, size = 6)) 
```
